Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-09 mr-review.Rmd
tweaks and added more to correlated horizontal pleio section

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable. The LD clumping procedure then selects SNPs that are not in high LD 
with the top-ranking SNPs.

LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by choosing a noncausal SNP that is in high LD with a 
causal SNP.

@@

Choosing a subset of the available SNPs to serve as instruments is one consideration that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs. However, this approach can be problematic when there are multiple causal 
SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy [@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] mdoels two varieties of horizontal pleiotropy:

>>>>

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable. The LD clumping procedure then selects high ranking SNPs that are not in high LD with each other.

LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs.

@@

Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
MRMix uses ...
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter. Specifically, CAUSE calculates the expected log pointwise posterior density 
between the two models. 
MRAID models horizontal pleiotropy by ...
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values.
